We might think about starting out this way.

Let's suppose that you are working with some other people and
collecting data by asking questions to people around town,
i.e. polling.  As you collect the information, you enter the results
into a spreadsheet for simplicity.  Then, when you are finished an
interview or just periodically, you email your results to your
colleagues.  Somebody, say Mary, is in charge of collecting the
worksheets from everyone and merging them into one and then sending
the unified sheet out to everyone.  This is a lot of work.  If
everyone is well-behaved and organized, this is simply just a lot of
tedious work for that person merging the sheets.  To make her job
slightly easier, she may require that you send only new data so that
she doesn't have to filter out repeats from previous submissions.  But
if somebody sends an old sheet, then problems may arise and a lot more
work may be needed to be done to recover from the error.  Ideally, the
data would be validated before being merged into the ``master''
sheet. And of course, before attempting the merge, a copy of the
original master sheet would be stored so that any errors could be
reversed and we could start over again from the last known correct
state.

Mary must also check that the mail came from people she knows and
not from somebody who happened to know she was gather this data
and who decided to send her some ``fake'' data.
So she must validate the submission.

Spreadsheets are quite simple to use, but it is also very easy to
enter a number as a string or vice-verse. Or perhaps somebody misses a
question and puts the answer from question 7 into column 6. So, Mary
now needs to check the types of the values in each column.  And not
necessarily just the type, but also whether the value is realistic or
at least sensible.  For example, if we get a value for age that is
less than $18$ when polling voters, we might be concerned.



Generally, we have issues of synchronization when merging across the
different data sources, validation of the data ``providers'', and
verification of the data types and values.
If one person cannot send or receive the updated spreadsheet, there is
an issue. 

Electronic voting

Manufacturing process with data being send to a centralized database
from many different devices.

Randomized clinical trial at many different hospitals.
Must coordinate the UUIDs for a patient




%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
