\documentclass{book}
\usepackage{comment}
\usepackage{times}
\usepackage{natbib}
\usepackage{fullpage}

\bibliographystyle{plain} % prsty, unsrt
%\usepackage[gather]{chapterbib}
\usepackage{bibunits}


\begin{document}

\begin{comment}
This is the  basic and initial outline
of topics we discussed on Saturday, 26th of June.
It is not in any way fixed.
\end{comment}

\tableofcontents


\include{preface}

\part{Programming}

\chapter{Introduction}
What is Statistical Computing, technology and informatics.

This is a reference \cite{GreenBook}.

\chapter{Introduction to Computing}
\section{Computation in General}
  Limitations of computers.
  Representing numbers,
  precision and comparison
  data types
  algorithms

  computing issues for statistics
    pivoting

   
  Programming
    control flow, data structures, etc.
    R, Matlab, Perl, Python, Java, C/C++, SAS, Excel

  Taxonomy of languages and systems.

\chapter{Programming Philosophy}

How to approach problems, writing functions, etc.  making code
general, extensible, etc.  Why is it important.  Thinking about
software as a library or inventory for use in future problems.

Advanced R (and other languages)

\include{Programming/thinking}

\chapter{Tools}

Analyzing particular problems and case studies and working out which
tools to use and how to combine them.

Think of an example of generating a report as an HTML page each night
by accessing data and fitting a model, etc.
shells, batch jobs, ssh, web servers, etc.

\chapter{Data/Case Studies}
Enumerate the different contexts
for the data we will use throughout the book.
This is a collection of unrelated sections
that give a background to e.g. TCP, email.

\include{Data/data}

\part{Technologies}

\chapter{RDBMS}
stored procedure, RinPostgres
\include{RDBMS/dbms}


\chapter{DCOM}

\chapter{Regular Expressions} 

\chapter{XML}

\chapter{GUIs}

\chapter{Case Studies}
extended/in depth

\chapter{Visualization}

\chapter{GIS}

\chapter{Reporting}

\chapter{HTTP and Sockets}

\chapter{SOAP}


\chapter{Inter-System Intefaces}
SJava, RSPerl, etc.

KEGG, PubMed


%\cite*

\cite{GreenBook}


\chapter{Bibliography}
\bibliographystyle{plain}
\bibliography{statcomp}



%%%\putbib[statcomp]

\end{document}
