\documentclass[11pt,timesroman]{article}
\usepackage{times}
\usepackage{fullpage}

\begin{document}

\begin{enumerate}

\item{Introduction - needs polish}
\begin{itemize}
        \item Data - three examples: rainfall, spam, traffic video
        \item Statistical Thinking and the Data Analysis process
        \item The role of computing in statistics and data analysis.
        \item Organization of rest of book (missing)
\end{itemize}

\paragraph{PART I: Technologies}

\item{Introduction to Computing - rough draft}
        \begin{itemize}
        \item History of the Pc
        \item Parts of the computer
        \item Taxonomy of languages and systems -- R, Matlab, Perl, Python, Java, C/C++, SAS, Excel
        \item Computation -- expressions, functions, variables 
        \item Bits, numbers, and characters
        \item Files and directories (class notes)
        \end{itemize}

\item{Shell Tools - detailed class notes}
        \begin{itemize}
        \item Command shell versus GUI
        \item File Management
        \item Pattern matching
        \item Process management 
        \item Advanced shell tools - scripts, looping, sed,
        \end{itemize}

\item{Introduction to Programming - detailed class notes}
        \begin{itemize}
        \item Expressions - single, compound, ill-formed
        \item Data types and structures, managing variables
        \item Functions and packages - help, arguments, scope, search path
        \item Operators - vectorized, recycling
        \item Subsetting - position, logical, exclusion, name
        \item Matrices and arrays
        \item Lists and data frames
        \end{itemize}

\item{Programming - detailed class notes}
% Connected to chapter Advanced programming.
        \begin{itemize}
        \item Loops, conditional statements, \texttt{ifelse}.
        \item Writing functions - 
                \begin{itemize}
                \item How to approach problems,
                \item General, Extensible
                \item Comments
                \end{itemize}
        \item Recursion.
        \item closures.
        \item Lazy evaluation.
        \item Dirty Dozen
        \end{itemize}


\item{Regular Expressions - class notes} 

\item{RDBMS - needs polish}
        \begin{itemize}
        \item Introduction 
        \item The basic relational model - the table
        \item Entity, meta information, missing values
        \item Queries and the Select statement
        \item Multiple tables and the relational model
        \item Interfaces - Accessing a database from R 
        \item SQL for statisticians - where to perform the statistics
        \item Considerations for managing and designing your own database
        \item Alternatives to databases
        \end{itemize}


\item{XML - needs polish}

        \begin{itemize}
        \item what is XML?
        \item XML syntax
        \item Valid XML 
        \item XML examples - OIM (SAS/Oracle/Microsoft), StatDataML, 
        \item The XML document as a tree
        \item Reading XML as a tree
        \item DOM Parser
        \item Walking the DOM object
        \item Customizing the DOM parser
        \item Examples - HTML, GML, 
        \item SAX
        \item Monitoring SAX.
        \item Hybrid parser model. 
        \end{itemize}

\item{SOAP - ?}

\item{Inter-System Interfaces - ?}
SJava, RSPerl, etc.
\\
KEGG, PubMed
\\
DCOM, GUIs, GIS, HTTP and Sockets


\item{Distributed Computing}
  \begin{itemize}
  \item HTTP requests and RCurl.
  \item Web Services and SOAP.
  \item DCOM.
  \item High-level view of SNOW.
  \item Inter-system interfaces. \\
      SJava, RSPerl, RMatlab, etc. \\
      GUIs 
  \end{itemize}
>>>>>>> 1.3

\item{Advanced computing - detailed class notes}
<<<<<<< toc_Nov_2005.tex
        \begin{itemize}
        \item Representing numbers, precision and comparison
        \item Efficiency - memory, time, profiling
        \end{itemize}
=======
        \begin{itemize}
        \item Representing numbers, precision and comparison
        \item Efficiency - memory, time, profiling
        \item Batch jobs.
        \item Debugging
        \item Classes - S3 and S4.
        \end{itemize}
>>>>>>> 1.3


\paragraph{Part II: \textit{Statistical} Computing - omit?}

\item{Simulation}
        \begin{itemize}
        \item Random number generation
        \item Inverse CDF, Acceptance rejection sampling
        \item Monte Carlo
        \item Resampling - bootstrap, cross-validation
        \item Monte Carlo Markov Chain
        \end{itemize}


\item{Visualization class notes}
<<<<<<< toc_Nov_2005.tex
        \begin{itemize}
        \item Elements of a graph
        \item eda
        \item complex composite plots
        \item Wilkinson's grammar of graphics
        \item Examples of complex graphs
        \item Dynamic graphics, Ggobi
        \end{itemize}
=======
        \begin{itemize}
        \item Elements of a graph
        \item eda
        \item complex composite plots
        \item Wilkinson's grammar of graphics
        \item Examples of complex graphs
        \item lattice.
        \item Dynamic graphics, Ggobi
        \end{itemize}
>>>>>>> 1.3


\paragraph{Part III: Case Studies}


\item{Spam - Naive Bayes (needs polish)}
\\
technology: regular expressions
\\
statistical methods: Naive Bayes; Cross-validation

\item{Spam - Nearest neighbor (needs polish)}
\\
technology: regular expressions
\\
statistical methods: nearest neighbor, classification trees,
cross-validation

\item{Election Maps - half finished}
\\
technology: understanding packages, gml
\\
statistical methods: maps, color


\item{Ad hoc Networks - detailed class notes}
\\
technology: random number generation and simulation
\\
statistical methods: acceptance-rejection sampling, monte carlo


\item{Google}
\\
technology: XML
\\
statistical methods: markov chains and simulation

\item{Baseball?}
\\
technology: databases
\\
statistical methods: exploratory data analysis


\item{Mailing Lists}
social networks


\item{Image analysis}

\item{Web Logs}

\item{TCP Intrusion detection}

\item{Genetics}
        SOAP

\item{Rainfall data and Geographical data.}
\item{Code Profiling}

\item{Citation clustering}


\item{Credit scoring}


\end{enumerate}

\appendix

\section{Getting, Installing and Working with R}
Installation.

Editors.


\section{Tools for Windows}

cygwin
XEmacs.

\end{document}
