\documentclass[11pt,timesroman]{article}
\usepackage{times}
\usepackage{fullpage}
\usepackage{comment}

\begin{document}
\begin{comment}
Other data possibilities -\\
thousands of satellite images of the ocean around the Farallon islands - used
in the search for Jim Gray's missing sailboat
\\
NASA registry of large pieces of junk in orbit around the earth - used
to plan space shuttle and satellite routes
\\
netcdf file - simulated data from weather models
\\
\end{comment}
\section*{Data}

\begin{tabular}{l|l|l|l}
Example & Introductory & Extended & Methods and Concepts\\
\hline
\hline
Satellite Image & Intro & & \\
Frontrange & Intro + Simple & & Missing data, Data in time \\
Spam & Intro + Reg Expr & Stand alone &Naive Bayes \\
Campaign donations & Digital & Digital &  \\
MP3 & Digital & &  \\
Aerial photo & Digital &  & \\
Traffic & Simple & Simple & \\
Super Nova & Simple & &SVM;  \\
Web Graph & Viz + XML & &\\
Geolocation & Viz & Viz & linear models and lattice \\
Election Map & Viz + XML & Viz & maps and packages \\
Dyanmic  & Viz & & \\
Hospitalization & Shell & & \\
NASA weather & Shell & Shell & \\
Weblog & Shell + RegExpr & Shell & data analysis using shell tools \\
Expt Design in profiling & Adv Pgm & & \\
Repeated Measures & Adv Pgm & & \\
State of Union & Reg Expr & Reg Expr & \\
Baseball & DB & DB & many to many matches \\
TCP & DB & & size \\
Census & DB & & \\
Google earth & XML & XML & \\
\end{tabular}

\section{Introduction}
  \begin{itemize}
  \item Philosophy - exploration; visualization is one of the most
    important tools of a statistician;
  \item Data are not just rectangular arrays of numbers - they need
    not be rectangular and they need not be numbers - three examples:
    rainfall, spam, traffic video
  \item Statistical Thinking and the Data Analysis process; the closer
    you are to orignial data the better job you will do at data
    analysis;
  \item The role of computing in statistics and data analysis.
    Empower student to explore
  \item Organization of rest of book
  \end{itemize}


\section{Digital data}
        \begin{itemize}
        \item Introductory examples
          \begin{itemize}
          \item Text numbers - Federal election campaign donations
          \item Music - MP3 song as a time series
          \item Image - Crowd in San Francisco demonstration
          \end{itemize}
        \item History of the PC
        \item Parts of the computer
        \item Taxonomy of languages and systems -- R, Matlab, Perl,
          Python, Java, C/C++, SAS, Excel
        \item Computation -- expressions, functions, variables 
        \item Random number generation
        \item Bits, numbers, and characters
        \item Files and directories
        \item Extended Example - federal election campaign donation
        \end{itemize}

\section{Exploring data in simple structures through R}
        \begin{itemize}
        \item Introductory Examples
          \begin{itemize}
            \item Traffic data
            \item Colorado Front range data
            \item Super nova data
          \end{itemize}
        \item Expressions - single, compound, ill-formed
        \item Data types and structures, managing variables
        \item Functions and packages - help, arguments, scope, search path
        \item Operators - vectorized, recycling
        \item Subsetting - position, logical, exclusion, name
        \item Matrices and arrays
        \item Lists and data frames
        \item read.table, scan, readlines
        \item Extended example - traffic data
        \end{itemize}


\section{An Introduction to Data Visualization}
        \begin{itemize}
          \item Introductory examples
            \begin{itemize}
              \item Lattice example - geolocation (repeat)?
              \item Social network graph - website network (repeat)?
              \item Dynamic graphic/video ?
              \item Map ? (election - repeat)
            \end{itemize}
        \item Elements of a graph
        \item eda
        \item complex composite plots
        \item Wilkinson's grammar of graphics
        \item Examples of complex graphs
        \item lattice.
        \item Dynamic graphics, Ggobi
        \item Extended example - lattice
        \item Extended example - election map
        \end{itemize}


\section{Simulating data \& An introduction to programming concepts}
        \begin{itemize}
          \item Introductory Examples
            \begin{itemize}
              \item Wireless - geolocation (cross-validation)
              \item Adhoc network
              \item Traffic congestion
            \end{itemize}
        \item Loops, conditional statements, \texttt{ifelse}.
          \\Acceptance rejection sampling
        \item Writing functions - 
                \begin{itemize}
                \item How to approach problems,
                \item General, Extensible
                \item Comments
                \item Inverse CDF
                \end{itemize}
        \item Recursion
        \item Closure
        \item Lazy evaluation.
        \item Resampling - bootstrap, cross-validation
        \item Monte Carlo Markov Chain
        \item Extended example - geolocation
        \item Extended example - Monte Carlo of adhoc network
        \end{itemize}


\section{Managing your data analysis with shell tools}
        \begin{itemize}
          \item Introductory examples
            \begin{itemize}
            \item Hospitalizations - johns hopkins data
            \item NASA weather data from competition
            \item Weblog - repeat
            \end{itemize}
          \item Command shell versus GUI
          \item File Management
          \item Pattern matching
          \item Process management
          \item Advanced shell tools - scripts, looping, sed,
          \item Extended example - NASA weather data (may be covered
            in previous sections already
          \item Extended example - weblog (data analysis using shell tools)
          \end{itemize}


\section{Advanced programming - separate chapter?}
        \begin{itemize}
        \item Introductory Examples
          \begin{itemize}
          \item Duncan's example of design of experiment in profiling?
          \item Duncan's example of repeated measures and errors?
          \end{itemize}
        \item Representing numbers, precision and comparison
        \item Efficiency - memory, time, profiling
        \item Batch jobs.
        \item Debugging
        \item Classes - S3 and S4.
        \item Extended example - ?
        \end{itemize}



\section{Handling free-form text data with regular expressions.} 
        \begin{itemize}
        \item Introductory examples
          \begin{itemize}
            \item Weblogs
            \item State of the union addresses
            \item Spam
          \end{itemize}
        \item literal strings 
        \item character classes 
        \item meta characters
        \item grouping and references 
 	\item alternation 
 	\item greedy matching
        \item Extended example - State of the union addresses
        \end{itemize}


\section{Data in Relational Databases}
        \begin{itemize}
        \item Introductory examples
          \begin{itemize}
            \item Baseball
            \item TCP intrusion detection
            \item Census data - ?
          \end{itemize}
        \item The basic relational model - the table
        \item Entity, meta information, missing values
        \item Queries and the Select statement
        \item Multiple tables and the relational model
        \item Interfaces - Accessing a database from R 
        \item SQL for statisticians - where to perform the statistics
        \item Considerations for managing and designing your own database
        \item Alternatives to databases
        \item Extended example - baseball
        \end{itemize}


\section{Self describing data - XML}
        \begin{itemize}
        \item Introductory examples
          \begin{itemize}
            \item Google earth - seal migration
            \item Something similar to Google's pagerank
            \item Election maps
          \end{itemize}
        \item what is XML?
        \item XML syntax
        \item Valid XML 
        \item XML examples - OIM (SAS/Oracle/Microsoft), StatDataML, 
        \item The XML document as a tree
        \item Reading XML as a tree
        \item DOM Parser
        \item Walking the DOM object
        \item Customizing the DOM parser
        \item Examples - HTML, GML, 
        \item SAX
        \item Monitoring SAX.
        \item Hybrid parser model. 
        \item Extended example - 
        \end{itemize}


\section{Inter-System Interfaces - ?}
SJava, RSPerl, etc.
\\
KEGG, PubMed
\\
DCOM, GUIs, GIS, HTTP and Sockets


\section{Data on the Internet ?}
  \begin{itemize}
  \item Introductory examples
    \begin{itemize}
      \item
    \end{itemize}
  \item HTTP requests and RCurl.
  \item Web Services and SOAP.
      GUIs 
  \end{itemize}



\section{Spam - Naive Bayes or Nearest neighbor}
technology: regular expressions
\\
statistical methods: Naive Bayes; Cross-validation
\\
statistical methods: nearest neighbor, classification trees,
cross-validation



\appendix

\section{Getting, Installing and Working with R}
Installation.

Editors.

\section{setting up a relational database}
SQL lite as an alternative

\section{Tools for Windows}

cygwin
XEmacs.

\end{document}
